High-accuracy neutron diffusion calculations based on integral transport theory

### 2019-12-11 docs/EPJP2019/Gross-EPJP-2019.tex
Update Gross-EPJP-2019.tex
Updated flow chart and description.
Some text and newcommants update.

\usetikzlibrary{patterns}
\usetikzlibrary{shapes,arrows}
\tikzstyle{discre} = [rectangle, draw, fill=white!20, 
text width=6em, text centered, minimum height=3em, node distance=3.5cm, line width=1pt]
\tikzstyle{decision} = [diamond, draw, fill=white!20, 
text width=3em, text badly centered, node distance=2cm, inner sep=1pt, line width=1pt]
\tikzstyle{block} = [rectangle, draw, fill=white!20, 
text width=6em, text centered, minimum height=3em, node distance=3.5cm, line width=1pt]
\tikzstyle{blockEq} = [rectangle, draw, fill=white!20, 
text width=8em, text centered, minimum height=3em, node distance=3.5cm, line width=0pt]
\tikzstyle{cloud} = [draw, ellipse,fill=white!20, node distance=4cm,
text centered, text width=3em, minimum height=2em, line width=1pt]
\tikzstyle{line} = [draw, -{latex'}, line width=1.0pt]

@@

\newcommand{\keff}{{\ensuremath{k_{\textrm{\scriptsize{eff}}}}}}
\newcommand{\beff}{\ensuremath{\beta_{\textrm{eff}}}}
\newcommand{\rr}{\ensuremath{\bm{r}}}
\newcommand{\OO}{\ensuremath{\hat{\bm{\Omega}}}}
\newcommand{\bnabla}{\ensuremath{\bm{\nabla}}}
\newcommand{\rE}{\ensuremath{(\rr,E)}}

\newcommand{\ftr}{\ensuremath{\phi_{\textrm{\scriptsize{tr}}}}}
\newcommand{\jtr}{\ensuremath{\bm{J}_{\textrm{\scriptsize{tr}}}}}
\newcommand{\jtrr}{\ensuremath{J_{\textrm{\scriptsize{tr}}}}}
%\newcommand{\mcL}{\mathcal{L}}
\newcommand{\mcL}{\tau}
\newcommand{\pl}[1]{\ensuremath{P_l(#1)}}
\newcommand{\dx}{\ensuremath{\Delta x}}

@@

Instead of computing new diffusion coefficients at the interfaces by Fick's law, $J_D(x,E) = -D(E) \partial_x \phi(x,E)$, new corrective currents $\delta J(x_s,E) = \jtrr(x_s,E) - J_D(x_s,E)$ are obtained at cell interfaces $x_s$. Here, $J_D(x,E)$ is called the ``diffusion current'' and is obtained using Fick's law, with the original values of the diffusion coefficients and with the derivative approximated by finite differences. In one-dimensional 
geometry and using the notation in \fig{fig:mesh1D}, the diffusion current and the corrective current are evaluated according to
\begin{subequations}\label{eq:JD-dJ}
\begin{align}
J_D(x_{i+1/2},E) \cong& -D(x_{i+1/2},E) 
\frac{\phi(x_{i+1},E) - \phi(x_i,E)}
{(\Delta x_{i+1} + \Delta x_i)/2}, \label{eq:JD} \\
\delta J(x_{i+1/2},E) =& -\delta D(x_{i+1/2},E) 
\frac{\phi(x_{i+1},E) + \phi(x_i,E)}
{(\Delta x_{i+1} + \Delta x_i)/2}, \label{eq:dJ}
\end{align}
\end{subequations}
where integer and rational subscripts indicate node-averaged and interface quantities, respectively. The discretized form of the current $\delta J$ must involve the neighboring flux as well, but its representation is changed into a drift-advection term to get rid of possible undefined division by zeros in case of flat flux~\cite{Smith-1983,Tomatis-2011}. The so-called ``transport current'' $\jtrr(x_s,E)$ is evaluated using the integral expression derived in~\sxn{sec:int-flux-J} as described in~\sxn{sec:calc-int-curr}.

@@

The new numerical corrections $\delta D$ are obtained at the interfaces using~\eqs{eq:dJ} to be used in the finite differences solver, together with the diffusive currents from~\eq{eq:JD}. Hence, the neutron balance resolved by the CMFD takes into account both types of currents $J_D$ and $\delta J$. %Non-linear iterations with new corrections given by $\delta J$ are needed because of their dependence on the unknown flux.
In particular, the one-dimensional multigroup neutron balance CMFD diffusion equations actually solved are
\begin{equation}\label{eq:diff1}
J_{D,g}^+(x) + \delta J_{D,g}^+(x) -  J_{D,g}^-(x) - \delta J_{D,g}^-(x) + \sigma_{g}\phi_g(x) =  \frac{\chi_g}{\keff}\sum\limits_{g'=1}^G\nu\sigma_{f,g'}(x)\phi_{g'}(x) + \sum\limits_{g'=1}^G\sigma_{s,g\leftarrow g'}\phi_{g'}(x),
\end{equation}
where $J_{D,g}^\pm (x)\equiv J_{D,g}(x_{i\pm 1/2})$.

@@

%
% ---------------------------------------------------
\subsection{Boundary conditions}
\label{sec:BC}
A generalized form for the boundary condition (at the left) follows as
\begin{equation}\label{eq:BC}
J_D(x=0) = -D_0\frac{\phi_0}{\Delta x_0/2 + \zeta},
\end{equation} 
where $\zeta$ is the extrapolation length in case of vacuum. The Marshak boundary conditions are reproduced by $\zeta=2D$, but the more accurate value $\zeta\approx 2.13D$ is taken to match transport with better agreement~\cite{Lamarsh-1966}. Reflection can be reproduced by $\zeta\rightarrow\infty$, whereas the condition of zero-flux is realized by $\zeta=0$. 
%
The quantity $\delta J$ at the boundary takes the simpler form of $\delta J = -\delta D_{-1/2}\phi_0$, without dividing by the spatial width, since no particular extrapolation length is appropriate for the correction. The expression for the right boundary is straightforward, implying a non-negative current. One should recall that the diffusion coefficient remains constant throughout the calculation, and so is the extrapolated distance. Only $\delta D$ is recalculated and the correction is actually implemented through the currents, i.e., $J_{tr} = J_D + \delta J$.

@@

The Ronen algorithm is described as follows (see Fig.~\ref{fig:RM-algo}):
\begin{enumerate}
	\item\label{it:init} Initialize - the algorithm receives the geometry and the cross sections (XS) of the problem as input. At this stage, the corrections to the diffusion coefficients $\delta D^{(0)}$ are set to zero and the initial pure-diffusion operator $\mathcal{A}_0$ is constructed. Initial guesses for the flux $\phi^{(0)}$ and multiplication factor $k^{(0)}_\texttt{eff}$ are set.
	\item\label{it:DF-solve} Diffusion Solver - a one-dimensional multigroup diffusion solver is executed until convergence using the diffusion corrections from the previous iteration $\delta D^{(k)}$ through the diffusion operator $\mathcal{A}^{(k)}=\mathcal{A}_0+\delta A^{(k)}$. The results are an updated estimation of the diffusion flux $\phi^{(k)}$ and the multiplication factor $k^{(k)}_\texttt{eff}$.
	\item\label{it:Js} Calculate Currents - using the updated diffusion flux $\phi^{(k)}$, both the diffusion current $J^{(k)}_D$ (Eq.~\ref{eq:JD-dJ}) and the transport current $J^{(k)}_\texttt{tr}$ (Eq.~\ref{eq:tot-J}) are calculated.
	\item\label{it:dD} Calculate corrections - the diffusion correction $\delta D^{(k+1)}$ (Eq.~\ref{eq:JD-dJ}) is obtained using $\delta J^{(k)}$ and Eq.~\eqref{eq:JD-dJ}.
	\item\label{it:dA} Update diffusion coefficients - reconstruct the diffusion operator $\mathcal{A}^{(k+1)} = \mathcal{A}_0+\delta \mathcal{A}^{(k+1)}$ (Eq.~\ref{eq:FD_diff_org}).
	\item\label{it:conv} If flux and eigenvalue are not converged, go to step~\ref{it:DF-solve}.
\end{enumerate}

\begin{figure}[htbp!] 
	\centering
\begin{tikzpicture}[font=\scriptsize]
% Place nodes
	\node [cloud] 					 			(start) {START};
	\node [block, right of=start, xshift=0cm,yshift=0cm] (input) {Input\\parameters};
	\node [block, right of=input, xshift=0cm,yshift=0cm] (geoXS) {Geometry and XS};
	\node [block, below of=geoXS, xshift=0cm,yshift=1.5cm] (Diff)  {Diffusion\\solver};
	\node [block, below of=Diff, xshift=0cm,yshift=1.5cm] (J)  {Calculate $J^{(k)}_{\texttt{tr}},J^{(k)}_D,\delta J^{(k)}$};
	\node [block, left  of=J,    xshift=0cm,yshift=0cm] (dD) {Obtain\\$\delta D^{(k+1)}$};
	\node [block, left  of=dD,   xshift=0cm,yshift=0cm](dA) {Calculate\\$\delta \mathcal{A}^{(k+1)}$};
	\node [block, above of=dA,   xshift=0cm,yshift=-1.5cm] (Anew) {Construct\\$\mathcal{A}_0+\delta \mathcal{A}^{(k+1)}$};

% Draw edges
	\path [line] (start) -- (input);
	\path [line] (input) -- (geoXS);
	\path [line] (geoXS) -- node[anchor= south west,yshift=-0.2cm]{$\mathcal{A}_0$}(Diff);
	\path [line] (Diff)  -- node[anchor= south west,yshift=-0.2cm]{$\phi^{(k)},k^{(k)}$}(J);
	\path [line] (J)  -- node[anchor= north]{}(dD);
	\path [line] (dD) -- node[anchor= north]{}(dA);
	\path [line] (dA)  -- node[anchor= north]{}(Anew);
	\path [line] (Anew)  -- node[near end,anchor= south west,xshift=-1.7cm]{$\mathcal{A}^{(k+1)}$}(Diff);

@@

Table~\ref{tab:dx_drho} shows the reactivity differences as a function of mesh refinement using standard diffusion (with extrapolated length) and the Ronen method. The reactivity difference is given by $(\Delta\rho = 1/k_\texttt{ref} - 1/k_\texttt{D/RM})\times 10^5$, where $k_D$ and $k_{RM}$ are the multiplication factors of standard diffusion and of the Ronen method, respectively. The reference $k_\texttt{ref}$ value from S\tsub{16} is 0.744417 ($I = 400$).

\begin{table}[htbp]
	\centering
	\caption{Reactivity differences as a function of mesh refinement using the Ronen method.}
	\label{tab:dx_drho}
	\begin{tabular}{lcccccr}
		%\midrule
		$dx$[cm] & $I$ &  $k_{\texttt{ref}}$ & $k_{\texttt{D}}$ & $\Delta\rho_\texttt{D}$ [pcm] & $k_{\texttt{RM}}$ & $\Delta\rho_\texttt{RM}$ [pcm]\\ 
		\midrule
		0.43 & ~50     & 0.744307 & 0.741417 & -524 & 0.740552  & -681\\
		%\midrule
		0.215 & 100    & 0.744391 & 0.741355 & -550 & 0.743447 & -171\\
		%\midrule
		0.1075 & 200   & 0.744412 & 0.741339 & -557 & 0.744212 & -36\\
		%\midrule
		0.07167 & 300 & 0.744416 & 0.741336 & -558 & 0.744356 & -11\\
		%\midrule
		0.05375 & 400  & 0.744417 & 0.741335 & -558 & 0.744407 & -2\\
		%\midrule
	\end{tabular}
\end{table}

@@

Numerically, Tomatis and Dall'Osso 2011 chose to deal with this potential singularity by using the well-known drift terms, which serve as a \emph{numerical feature}. The (more accurate) surface current $J_{\texttt{tr}}$, calculated by integral expression, is written as the sum of the surface current calculated by diffusion, $J_\texttt{D}$, plus a correction $\delta J$. Hence, when the current vanishes, the correction $\delta J$ vanishes as well. Moreover, the correction is written as in Eq.~\eqref{eq:JD-dJ}, avoiding the potential divergence resulting from division by zero.

>>>>

\usetikzlibrary{patterns}
\usetikzlibrary{shapes,arrows}
\tikzstyle{discre} = [rectangle, draw, fill=white!20, 
text width=6em, text centered, minimum height=3em, node distance=3.5cm, line width=1pt]
\tikzstyle{decision} = [diamond, draw, fill=white!20, 
text width=4em, text badly centered, node distance=2cm, inner sep=2pt, line width=1pt]
\tikzstyle{block} = [rectangle, draw, fill=white!20, 
text width=6em, text centered, minimum height=3em, node distance=3.5cm, line width=1pt]
\tikzstyle{blockEq} = [rectangle, draw, fill=white!20, 
text width=8em, text centered, minimum height=3em, node distance=3.5cm, line width=0pt]
\tikzstyle{cloud} = [draw, ellipse,fill=white!20, node distance=4cm,
text centered, text width=3em, minimum height=2em, line width=1pt]
\tikzstyle{line} = [draw, -{latex'}, line width=1.0pt]

@@

\newcommand{\keff}{{\ensuremath{k_{\textrm{\scriptsize{eff}}}}}}
\newcommand{\kref}{{\ensuremath{k_{\textrm{\scriptsize{ref}}}}}}
\newcommand{\krm}{{\ensuremath{k_{\textrm{\scriptsize{RM}}}}}}
\newcommand{\kd}{{\ensuremath{k_{\textrm{\scriptsize{D}}}}}}
\newcommand{\beff}{\ensuremath{\beta_{\textrm{eff}}}}
\newcommand{\rr}{\ensuremath{\bm{r}}}
\newcommand{\OO}{\ensuremath{\hat{\bm{\Omega}}}}
\newcommand{\bnabla}{\ensuremath{\bm{\nabla}}}
\newcommand{\rE}{\ensuremath{(\rr,E)}}

\newcommand{\ftr}{\ensuremath{\phi_{\textrm{\scriptsize{tr}}}}}
\newcommand{\jtr}{\ensuremath{\bm{J}_{\textrm{\scriptsize{tr}}}}}
\newcommand{\jtrr}{\ensuremath{J_{\textrm{\scriptsize{tr}}}}}
\newcommand{\jd}{\ensuremath{J_{\textrm{\scriptsize{D}}}}}
\newcommand{\delj}{\ensuremath{\delta J}}
%\newcommand{\mcL}{\mathcal{L}}
\newcommand{\mcL}{\tau}
\newcommand{\pl}[1]{\ensuremath{P_l(#1)}}
\newcommand{\dx}{\ensuremath{\Delta x}}

@@

Instead of computing new diffusion coefficients at the interfaces by Fick's law, $\jd(x,E) = -D(E) \partial_x \phi(x,E)$, new corrective currents $\jd(x_s,E) = \jtrr(x_s,E) - \jd(x_s,E)$ are obtained at cell interfaces $x_s$. Here, $\jd(x,E)$ is called the ``diffusion current'' and is obtained using Fick's law, with the original values of the diffusion coefficients and with the derivative approximated by finite differences. In one-dimensional 
geometry and using the notation in \fig{fig:mesh1D}, the diffusion current and the corrective current are evaluated according to
\begin{subequations}\label{eq:JD-dJ}
\begin{align}
\jd(x_{i+1/2},E) \cong& -D(x_{i+1/2},E) 
\frac{\phi(x_{i+1},E) - \phi(x_i,E)}
{(\Delta x_{i+1} + \Delta x_i)/2}, \label{eq:JD} \\
\delta J(x_{i+1/2},E) =& -\delta D(x_{i+1/2},E) 
\frac{\phi(x_{i+1},E) + \phi(x_i,E)}
{(\Delta x_{i+1} + \Delta x_i)/2}, \label{eq:dJ}
\end{align}
\end{subequations}
where integer and rational subscripts indicate node-averaged and interface quantities, respectively. The discretized form of the current $\delj$ must involve the neighboring flux as well, but its representation is changed into a drift-advection term to get rid of possible undefined division by zeros in case of flat flux~\cite{Smith-1983,Tomatis-2011}. The so-called ``transport current'' $\jtrr(x_s,E)$ is evaluated using the integral expression derived in~\sxn{sec:int-flux-J} as described in~\sxn{sec:calc-int-curr}.

@@

The new numerical corrections $\delta D$ are obtained at the interfaces using~\eqs{eq:dJ} to be used in the finite differences solver, together with the diffusive currents from~\eq{eq:JD}. Hence, the neutron balance resolved by the CMFD takes into account both types of currents $\jd$ and $\delj$. %Non-linear iterations with new corrections given by $\delj$ are needed because of their dependence on the unknown flux.
In particular, the one-dimensional multigroup neutron balance CMFD diffusion equations actually solved are
\begin{equation}\label{eq:diff1}
J_{D,g}^+(x) + \delj_g^+(x) -  J_{D,g}^-(x) - \delj_g^-(x) + \sigma_{g}\phi_g(x) =  \frac{\chi_g}{\keff}\sum\limits_{g'=1}^G\nu\sigma_{f,g'}(x)\phi_{g'}(x) + \sum\limits_{g'=1}^G\sigma_{s,g\leftarrow g'}\phi_{g'}(x),
\end{equation}
where $J_{D,g}^\pm (x)\equiv J_{D,g}(x_{i\pm 1/2})$.

@@

%
% ---------------------------------------------------
\subsection{Boundary conditions}
\label{sec:BC}
A generalized form for the boundary condition (at the left) follows as
\begin{equation}\label{eq:BC}
\jd(x=0) = -D_0\frac{\phi_0}{\Delta x_0/2 + \zeta},
\end{equation} 
where $\zeta$ is the extrapolation length in case of vacuum. The Marshak boundary conditions are reproduced by $\zeta=2D$, but the more accurate value $\zeta\approx 2.13D$ is taken to match transport with better agreement~\cite{Lamarsh-1966}. Reflection can be reproduced by $\zeta\rightarrow\infty$, whereas the condition of zero-flux is realized by $\zeta=0$. 
%
The quantity $\delj$ at the boundary takes the simpler form of $\delj = -\delta D_{-1/2}\phi_0$, without dividing by the spatial width, since no particular extrapolation length is appropriate for the correction. The expression for the right boundary is straightforward, implying a non-negative current. One should recall that the diffusion coefficient remains constant throughout the calculation, and so is the extrapolated distance. Only $\delta D$ is recalculated and the correction is actually implemented through the currents, i.e., $\jtrr = \jd + \delj$.

@@

The Ronen algorithm is described as follows (see Fig.~\ref{fig:RM-algo}):
\begin{enumerate}
	\item\label{it:init} Initialize - the algorithm receives the geometry and the cross sections (XS) of the problem as input. At this stage, the correction factors $\delta D^{(0)}$ are set to zero and the initial pure-diffusion operator $\mathcal{A}_0$ is constructed. Initial guesses for the flux $\phi^{(0)}$ and multiplication factor $\keff^{(0)}$ are set.
	\item\label{it:Js} Calculate currents - using the diffusion flux $\phi^{(k)}$, both the diffusion current $\jd^{(k)}$ (Eq.~\ref{eq:JD-dJ}) and the transport current $\jtrr^{(k)}$ (Eq.~\ref{eq:tot-J}) are calculated.
	\item\label{it:dD} Calculate correction factors - the correction factors $\delta D^{(k+1)}$ (Eq.~\ref{eq:JD-dJ}) is obtained using $\delj^{(k)}$ and Eq.~\eqref{eq:JD-dJ}.
	\item\label{it:dA} Update diffusion operator - reconstruct the diffusion operator $\mathcal{A}^{(k+1)} = \mathcal{A}_0+\delta \mathcal{A}^{(k+1)}$ (Eq.~\ref{eq:FD_diff_org}).
	\item\label{it:DF-solve} Diffusion solver - a one-dimensional multigroup diffusion solver is executed until convergence. The results are an updated estimation of the diffusion flux $\phi^{(k+1)}$ and the multiplication factor $\keff^{(k+1)}$.
	\item\label{it:cvg} Check convergence - check if the flux $\phi^{(k)}$ and the multiplication factor $\keff^{(k)}$ converged. If yes, stop, otherwise, step~\ref{it:Js}.
\end{enumerate}

\begin{figure}[htbp!] 
	\centering
\begin{tikzpicture}[font=\scriptsize]
% Place nodes
	\node [cloud] 					 						(start) {START};
	\node [block, below of=start, xshift=0cm,yshift=1.5cm]  (input) {Input\\parameters};
	\node [block, below of=input, xshift=0cm,yshift=1.5cm]  (geoXS) {Geometry, XS};
	\node [block, below of=geoXS, xshift=0cm,yshift=1.5cm]  (J)     {Calculate $\jtrr^{(k)},\jd^{(k)},\delj^{(k)}$};
	\node [block, below of=J,     xshift=0cm,yshift=1.5cm]  (dDA)   {Obtain $\delta D^{(k+1)}$, $\delta \mathcal{A}^{(k+1)}$};
	\node [block, below of=dDA,  xshift=0cm,yshift=1.5cm]  	(Anew)  {Construct\\$\mathcal{A}_0+\delta \mathcal{A}^{(k+1)}$};
	\node [block, below of=Anew, xshift=0cm,yshift=1.5cm]  	(Diff)  {Diffusion\\solver};
	\node [decision, below of=Diff,xshift=0cm,yshift=-1cm]	(cvg)   {$\phi^{(k+1)}$, $k^{(k+1)}$ \\ converged?};
	\node [cloud, below of=cvg,xshift=0cm,yshift=1.5cm] 	(stp)  {STOP};

% Draw edges
	\path [line] (start) -- (input);
	\path [line] (input) -- (geoXS);
	\path [line] (geoXS) -- node[anchor= south west,yshift=-0.2cm]{$\phi^{(0)}$, $k^{(0)}$, $\mathcal{A}_0$}(J);
	\path [line] (J) 	 -- (dDA);
	\path [line] (dDA) 	 -- (Anew);
	\path [line] (Anew)  -- node[near end,anchor= south west,xshift=0cm]{$\mathcal{A}^{(k+1)}$}(Diff);
	\path [line] (Diff)  -- node[anchor= south west,yshift=-0.2cm]{$\phi^{(k+1)},k^{(k+1)}$}(cvg);
	\path [line] (cvg)   -- node[anchor= south west]{YES}(stp);
	\path [line] (cvg)   -| node[near start,yshift=0.2cm]{NO}(-3,-6.0) -- node[near end]{}(J);

@@

Table~\ref{tab:dx_drho} shows the reactivity differences as a function of mesh refinement using standard diffusion (with extrapolated length) and the Ronen method. The reactivity difference is given by $(\Delta\rho = 1/\kref - 1/k_\texttt{D/RM})\times 10^5$, where $\kd$ and $\krm$ are the multiplication factors of standard diffusion and of the Ronen method, respectively. The reference $\kref$ value from S\tsub{16} is 0.744417 ($I = 400$).

\begin{table}[htbp]
	\centering
	\caption{Reactivity differences as a function of mesh refinement using the Ronen method.}
	\label{tab:dx_drho}
	\begin{tabular}{lcccccr}
		%\midrule
		$dx$[cm] & $I$ &  $\kref$ & $\kd$ & $\Delta\rho_\texttt{D}$ [pcm] & $\krm$ & $\Delta\rho_\texttt{RM}$ [pcm]\\ 
		\midrule
		0.43 & ~50     & 0.744307 & 0.741417 & -524 & 0.740552  & -681\\
		%\midrule
		0.215 & 100    & 0.744391 & 0.741355 & -550 & 0.743447 & -171\\
		%\midrule
		0.1075 & 200   & 0.744412 & 0.741339 & -557 & 0.744212 & -36\\
		%\midrule
		0.07167 & 300 & 0.744416 & 0.741336 & -558 & 0.744356 & -11\\
		%\midrule
		0.05375 & 400  & 0.744417 & 0.741335 & -558 & 0.744407 & -2\\
		%\midrule
	\end{tabular}
\end{table}

@@

Numerically, Tomatis and Dall'Osso 2011 chose to deal with this potential singularity by using the well-known drift terms, which serve as a \emph{numerical feature}. The (more accurate) surface current $\jtrr$, calculated by integral expression, is written as the sum of the surface current calculated by diffusion, $\jd$, plus a correction $\delj$. Hence, when the current vanishes, the correction $\delj$ vanishes as well. Moreover, the correction is written as in Eq.~\eqref{eq:JD-dJ}, avoiding the potential divergence resulting from division by zero.
